$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-04-30 paper/paper.tex
Go through the contributions section

The \picalc{} also benefited from this linearity wave in early '90s.
Kobayashi et al. \cite{KPT96} defined the {linear} \picalc{}, a typed version of the \picalc{} where the linear type system restricts the usage of channels, which must be used exactly once.

@@

\subsection{Contributions}
Our contributions are given in the following together with the corresponding sections.
\begin{itemize}
  \item \textbf{Machine verified formalisation of the syntax and semantics of the \picalc{}}: The syntax for the \picalc{} is given in \S \ref{syntax}.
  We use de Bruijn indices \cite{} as variable references, and dependent types to make the syntax well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  We define the operational semantics of the \picalc{} by a reduction relation in \S \ref{operational-semantics}.
  The base case of every such reduction is communication over a channel; we take care to track this information at the type level so that we can then use it for subject reduction (aka type preservation) theorem.

  \item \textbf{Machine verified typing with leftovers for the \picalc{} with linear, gradual and shared types}:
  Linear type systems need of context splits.
  These are tricky to handle constructively without explicitly asking the user to supply them.
  Leftover typing \cite{} is a type theoretical technique that addresses this.
  It does so by carrying both an input and an output context (the leftovers).
  The output context can then be used as the input context of the continuations.
  We apply this technique to the \picalc{} in \S \ref{leftover-typing}.
  We generalize our algebra on usage multiplicities in \S \ref{multiplicities}.
  As a result, we are able to model a mix of shared, linear, and gradual types.
  \todo{reformulate paragraph}
  Another property of particular interest in concurrent and distributed systems is the avoidance of race conditions.
  A type system that excludes multiple usages of a resource excludes race conditions. Linear \picalc{} not only avoids races, but ensures \emph{communication privacy}: if both endpoints of a channel are acquired no other process can eavesdrop.
  It also ensures that every declared communication channel must be used.
  We provide a typing system for the linear \picalc{} in \S \ref{type-system} as extensional rules on sitting on top of the unruly well-scoped syntax.

  \item \textbf{Machine verified proofs of subject reduction and auxiliary lemmas for all $\pi$ with leftovers}:

  \item \textbf{Full formalisation in Agda available at \cite{Zalakain2020Agda}}:
\end{itemize}

%\paragraph*{Syntax}
%In the \lambdacalc{} all computations are local and happen on syntactically contiguous terms; in the \picalc{} however communicating subprocesses can be non-contiguous and even have different parent processes.
%One way to define a reduction relation modulo these syntactic minutiae is to define a structural congruence \cite{} relation that acts as a quotient type.
%A term $A$ can then be freely rewritten into a term $B$ as long as $A$ is structurally congruent to $B$.

%\paragraph*{Semantics}
% In the \lambdacalc{} function application is the main computational engine; in the \picalc{} it is inter-process communication what drives computation.
% Two subprocesses can communicate anywhere in a process but under input and output, which must be respected to ensure the sequential behavior.
% We type a reduction relation that acts as an operational semantics modulo subject congruence in \S \ref{operational-semantics}.
% The base case of every such reduction is communication over a channel; we take care to track this information at the type level so that we can then use it to state the theorem for subject reduction (aka type preservation).

% As with the untyped \lambdacalc{}, the untyped \picalc{} can be used to construct a wide variety of ill-behaved terms --- for some characterization of ill-behaved.
% One can use type systems to constructively obtain a subset of terms for which certain properties hold.
% As an example, the simply typed \lambdacalc{} ensures that the argument supplied to a function is of the expected type.
% Correspondingly, the simply typed \picalc{} \cite{} ensures that the data sent over a channel is of the expected type.
% This property is often referred to as \emph{communication safety}.

% \todo{Mention that our semantics is prior to our typing (Curry-style)}

% Another property of particular interest to concurrent systems is the avoidance of race conditions.
% A type system that excludes multiple usages of a resource excludes race conditions.
% In the linear \picalc{} \cite{} channels must be used exactly once.
% This not only avoids races, but ensures \emph{communication privacy}: if both endpoints of a channel are acquired no other process can eavesdrop.
% It also ensures that every declared communication channel must be used.
% The linear \picalc{} is also interesting because together with product types it is isomorphic to session types \cite{} -- which are used to encode communication protocols.
% We provide a typing system for the linear \picalc{} in \S \ref{type-system} as extensional rules on sitting on top of the unruly well-scoped syntax.

% \paragraph*{Typing with leftovers}
% Linear type systems need of context splits.
% These are tricky to handle constructively without explicitly asking the user to supply them.
% Leftover typing \cite{} is a type theoretical technique that addresses this.
% It does so by carrying both an input and an output context (the leftovers).
% The output context can then be used as the input context of the continuations.
% We apply this technique to the \picalc{} in \S \ref{leftover-typing}.
% We generalize our algebra on usage multiplicities in \S \ref{multiplicities}.
% As a result, we are able to model a mix of shared, linear, and gradual types.
% \todo{reformulate paragraph}

@@

Data type definitions have double lines and index-free synonyms as rule names (for ease of reference).
We otherwise use the constructor name as the name of a typing rule.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructor names are overloaded --- they and disambiguated by context.

>>>>

The \picalc{} also benefited from this linearity wave in the early '90s.
Kobayashi et al. \cite{KPT96} defined the {linear} \picalc{}, a typed version of the \picalc{} where the linear type system restricts the usage of channels, which must be used exactly once.

@@

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and semantics of the \picalc{}}:
  Section \S \ref{syntax} uses type level de Bruijn indices \cite{} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  Section \S \ref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- ak-a type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unecessary syntactic minutiae introduced by the syntax of the \picalc{}.

  \item \textbf{Leftover typing for the \picalc{}, with linear, gradual and shared types}:
  Section \S \ref{type-system} provides the type system for a resource-aware \picalc{}.
  The type system is \emph{resource-aware} in that it allows the user to provide a resource-aware algebra that is then applied to the type-system.
  Linear types, gradual types and shared types are all instances of such a resource-aware algebra too.
  Any \emph{partial commutative monoid} that is cancellative and has a minimal element is a valid such algebra.
  Multiple of these algebras can be applied to a single type system --- contexts keep information about what algebra should be applied to which element.

  The typing judgments use \emph{leftover typing} to model the resource-aware \picalc{}.
  This approach adds a leftover usage context to the typing judgments.
  Typing derivations take the resources of their input usage context, consume some of them, and leave the rest as leftovers in the output usage context.
  These technique renders explicit context splits unecessary.

  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  Section \S \ref{type-safety} builds on several theorems to proof subject reduction for our type system.
  Here too, leftover typing renders the proofs for some of these lemmas straightforward, while for others it makes it possible to state them in the first place.
  We prove \textbf{framing} (well-typedness is independent of the unused resources of a process), \textbf{weakening} (new variables that are added to both the input and output usage context preserve well-typedness), and \textbf{strengthening} (removing the variables that are unused by a process preserves well-typedness).
  We use these to prove \textbf{subject congruence} (structural congruence rules can be freely applied while preserving well-typedness) and \textbf{substitution} (a variable can be substituted for another as long as this other variable is capable of doing everything the original one did), and use these to prove \textbf{subject reduction} (a well-typed process that reduces will still be well-typed).

  \item \textbf{Fully mechanised formalisation available in Agda}:
  This work has been fully formalised in Agda and is publicly avaible \cite{Zalakain2020Agda}.

\end{itemize}

@@

Data type definitions use double lines and index-free synonyms as rule names (for ease of reference).
We otherwise use the constructor name as the name of a typing rule.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructor names are overloaded --- they and disambiguated by context.
